Dependently Typed Multi-stage Calculus

### 2019-11-26 slides/main.tex
Finished third draft.

% Remove all navigation symbols
\setbeamertemplate{navigation symbols}{}
% using non standard fonts for beamer
\usefonttheme{professionalfonts}
% Set the font size of titles
% See other font category at
% https://tex.stackexchange.com/questions/183052/what-are-all-the-possible-first-arguments-to-setbeamerfont
\setbeamerfont{title}{size=\normalsize}
\setbeamerfont{frametitle}{size=\normalsize}
\setbeamerfont{normal text}{size=\small}
\setbeamerfont{itemize/enumerate body}{size=\small}
\setbeamerfont{itemize/enumerate subbody}{size=\footnotesize}
\setbeamerfont{frametitle continuation}{size=\small}
\setbeamerfont{framesubtitle}{size=\small}
\setbeamerfont{abstract}{size=\small}
\setbeamerfont{projected text}{size=\small}
\setbeamerfont{block title}{size=\small}
\setbeamerfont{footline}{size=\tiny}

@@

\begin{frame}[fragile]{Example of Multi-stage Programming in MetaOCaml}
    \begin{center}
        \begin{verbatim}
 # let a = < (1 + 2) >;;
 val a : int code = < (1 + 2) >
 # run a;;
 - : int = 3

 # let b = <(~a) + (~a)>
 val b : int code = < ((1 + 2) + (1 + 2)) >
 # run b
 - : int = 6
        \end{verbatim}
    \end{center}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Application of Multi-stage Programming: vadd-gen}
    Generate efficient vector addition functions for given length.

    \begin{verbatim}
 val vadd-gen : Int -> (Vec -> Vec -> Vec) code
 # let vadd3 = vadd-gen 3
 val vadd3 : (Vec -> Vec -> Vec) code
 (* vadd3 v w =                            *)
 (*   < let (v, w)=(~v, ~w) in             *)
 (*      [v[0]+w[0];v[1]+w[1];v[2]+w[2]] > *)
    \end{verbatim}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{The problem of vadd3 generated by vadd-gen}
    Runtime error occurs for vectors of wrong length.
    \begin{verbatim}
 val vadd3 : (Vec -> Vec -> Vec) code
 # (run vadd3) [1;2;3] [4;5;6]
 - : Vec = [5;7;9]
 # (run vadd3) [1;2] [4;5]
 Exception: Failure
    \end{verbatim}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Dependent Types}
    \renewcommand{\V}{\text{Vec}\ }
    Types dependent on values
    \begin{itemize}
        \item Can prevent illegal use of {\verb|vadd3|}.
    \end{itemize}
 \begin{verbatim}
 [1;2]: Vec 2
 vadd3 : (Vec 3 -> Vec 3 -> Vec 3) code

 # (run vadd3) [1;2] [4;5]
 [1;2] has type Vec 2
 but was expected of type Vec 3
 \end{verbatim}
    \note{
        By the way, we can write types dependent on values using dependent types.
    }
\end{frame}

@@

\begin{frame}[fragile]{\LTP [Hanada\&Igarashi'14]}
    Multi-stage calculus with Cross Stage Persistence (CSP)
    \begin{block}{Terms}
        \begin{tabbing}
            \hspace{5mm} \( M \) \= \( ::= c \) \hspace{20mm} \= (constants) \\
            \> \( \mid \TB M \) \> (quote) \\
            \> \( \mid \TBL M \) \> (unquote) \\
            \> \( \mid \textbf{run}\ M \) \> (run) \\
            \> \( \mid \% M \) \> (Cross Stage Persistence (CSP)) \\
            \> \( \mid x \mid \lambda x:\tau.M \mid M\ M \) \\
        \end{tabbing}
    \end{block}
\end{frame}

\begin{frame}[fragile]{Types of \LTP}
    \begin{block}{Types}
        \begin{tabbing}
            \hspace{5mm} \( \tau \) \= ::= X \hspace{20mm} \= (constants) \\
            \> \( \mid \TW\ \tau \) \> (code type) \\
        \> \( \mid \tau \to \tau \) \> (function type)
    \end{tabbing}
    \end{block}
   \note{
    }
\end{frame}

\begin{frame}[fragile]{Example of \LTP}
    \begin{table}
        \begin{tabular}{ c | c | c }
            & \LTP & MetaOCaml \\[2mm]
            \hline
            Quote & \( \textbf{let } a = \TB (1 + 2) \) & \verb| let a = .<(1+2)>.| \\[2mm]
            Unquote & \( (\TB (\TBL a + \TBL a)) \) & \verb| .<(.~a)+(.~a)>.| \\[2mm]
            Run & \( \textbf{run } (\TB (1 + 2)) \) & \verb| run .<(1+2)>.| \\[2mm]
            Code type & \( \TW\ \text{Int} \) & \verb| int code |
        \end{tabular}
    \end{table}
\end{frame}

\begin{frame}[fragile]{What is a stage?}
    Level of nested quote
    \begin{center}
        \( \underbrace{(\textbf{run}\ (\TB\ \overbrace{(2 + 3)}^{\let\scriptstyle\textstyle\substack{\text{stage }1}})) + 1}_{\let\scriptstyle\textstyle\substack{\text{stage }0}} \)
    \end{center}

@@

\begin{frame}[fragile]{Typing Rules of \LTP}
    Cannot apply a term to a function in a different stage.
    \begin{center}
        {\small \infrule[\TApp]{
            \G\V M:\sigma \to \tau @ n \andalso
            \G\V N:\sigma @ n
        }{
            \G\V M\ N : \tau @ n
        }}
    \end{center}
    Increase and decrease stage using \( \TB \) and \( \TBL \).
    \begin{center}
        {\small \infrule[{\TTB}]{
            \G\V M:\tau @ n+1
        }{
            \G\V\TB M:\TW \tau @ n
        } \hfil
        \infrule[{\TTBL}]{
            \G\V M:\TW \tau @ n
        }{
            \G\V\TBL M:\tau @ n + 1
        }}
    \end{center}
    In general, we cannot use terms of one stage at other stages.
    \note{
        Let's have a look on some typing rules.
    }
\end{frame}

@@

    Want to use \( \text{mul5} \) in a code value.
    \begin{center}
        \( \textbf{let}\ \underbrace{\text{mul5}}_{\text{stage }0}\ x = 5 * x\ \textbf{in} \TB\ \underbrace{(\% \text{mul5}\ 4)}_{\text{stage }1} \)
    \end{center}
    CSP symbol \( \% \) {\bf{lift up}} mul5 from stage 0 to stage 1.
\end{frame}

@@

\begin{frame}[fragile]{Kinding rules of \LMD}
    Types have also stages because types may contains terms.
    \begin{center}
    \infrule[{\KApp}]{
        \G\V \sigma:: (\Pi x:\tau.K) @ n \andalso \G\V M:\tau @ n
    }{
        \G\V \sigma\ M::K[x\mapsto M] @ n
    }
    \infrule[{\KAbs}]{
        \G\V \tau :: * @ n \andalso \G,x:\tau@ n \V \sigma::J@A
    }{
        \G\V(\Pi x:\tau.\sigma) :: (\Pi x:\tau.J) @ n
    }
    \end{center}
    So, we need CSP in types also.
    \begin{center}
    \( \underbrace{\Pi n:\I. \TW (\overbrace{\text{Vec}\ \%n}^{\text{stage }1})}_{\text{stage }0} \)
    \end{center}
\end{frame}

\begin{frame}[fragile]{Constants and type-level constants of \LMD}
    Constants and type-level constants can be used at any stage.
    \begin{itemize}
        \item Integers: \( 0,1,2,3,\dots \)
        \item Int, Vec, \dots
    \end{itemize}

    \begin{tabbing}
        \hspace{5mm} \= \text{val doub3} : \( \text{Vec }3 \to \text{Vec }3 \) \\[2mm]
        \> \textbf{let } \text{double} = \( \TB (\% \text{doub3}) \) \\
        \> \text{val double} : \( \TW ( \text{Vec }3 \to \text{Vec }3) \)
    \end{tabbing}
\end{frame}

\begin{frame}[fragile]{vadd-gen in \LMD}
    Type tells us the length of vector.
    \newcommand{\Vn}{\text{Vec}\ \%n}
    \newcommand{\Vt}{\text{Vec}\ \%3}
    \begin{tabbing}
        \hspace{5mm} \= \( \text{vadd-gen} : \Pi n:\I. \TW (\Vn \to \Vn \to \Vn) \) \\[2mm]
        \> \( \textbf{let}\ \text{vadd3} = \text{vadd-gen}\ 3;; \) \\
        \> \( \text{val}\ \text{vadd3} : \TW (\Vt \to \Vt \to \Vt) \) \\[2mm]
    \end{tabbing}
    Now, vadd3 takes only vectors of length 3.
    \note{
        vadd-gen is changed as follows in \LMD.
        Thanks to dependent types ...
    }
\end{frame}

\begin{frame}[fragile]{Problem of type of vadd3}
    Hard to compose with other functions.
    \newcommand{\Vn}{\text{Vec}\ \%n}
    \newcommand{\Vpt}{\text{Vec}\ \%3}
    \newcommand{\Vt}{\text{Vec}\ 3}
    \begin{tabbing}
        \hspace{5mm} \= \( \text{val}\ \text{vadd3} : \TW (\Vpt \to \Vpt \to \Vpt) \) \\
        \> \( \text{val}\ \text{double} : \TW (\Vt \to \Vt) \) \\[2mm]
        \> \( \textbf{let}\ \text{add-double}\ v\ w = \) \\
        \> \hspace{5mm} \( \TB (\underbrace{\TBL\text{double}}_{\Vt \to \Vt}\ \underbrace{(\TBL \text{vadd3}\ \TBL v\ \TBL w)}_{\Vpt});; \) \\
    \end{tabbing}
    We cannot compose because of type mismatch.
\end{frame}

\begin{frame}[fragile]{Solution}
    \newcommand{\Vt}{\text{Vec}\ 3}
    \newcommand{\Vpt}{\text{Vec}\ \%3}
    Erase the CSP symbol \( \% \) of an closed term.
    \begin{center}
        \infrule[{\QPercent}]{
            M \text{ is a closed term.}
        }{
            \G\vdash \% M \E M
        }
    \end{center}
    \begin{itemize}
        \item \( \text{Vec}\ \%3 \E \text{Vec}\ 3 \)
        \item \( \text{vadd3} : \TW (\Vt \to \Vt \to \Vt) \) \\
            \hspace{15mm} \( (\E  \TW (\Vpt \to \Vpt \to \Vpt)) \)
    \end{itemize}
\end{frame}

>>>>

% Remove all navigation symbols
\setbeamertemplate{navigation symbols}{}
% using non standard fonts for beamer
\usefonttheme{professionalfonts}
% Set the font size of titles
% See other font category at
% https://tex.stackexchange.com/questions/183052/what-are-all-the-possible-first-arguments-to-setbeamerfont
\setbeamerfont{title}{size=\normalsize}
\setbeamerfont{frametitle}{size=\normalsize}
\setbeamerfont{normal text}{size=\small}
\setbeamerfont{itemize/enumerate body}{size=\small}
\setbeamerfont{itemize/enumerate subbody}{size=\small}
\setbeamerfont{frametitle continuation}{size=\small}
\setbeamerfont{framesubtitle}{size=\small}
\setbeamerfont{abstract}{size=\small}
\setbeamerfont{projected text}{size=\small}
\setbeamerfont{block title}{size=\small}
\setbeamerfont{footline}{size=\tiny}

@@

\begin{frame}[fragile]{Example of Multi-stage Programming in MetaOCaml}
    \begin{exampleblock}{}
        \begin{verbatim}
 # let a = < (1 + 2) >;;
 val a : int code = < (1 + 2) >
 # run a;;
 - : int = 3

 # let b = <(~a) + (~a)>
 val b : int code = < ((1 + 2) + (1 + 2)) >
 # run b
 - : int = 6
        \end{verbatim}
    \end{exampleblock}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Application of Multi-stage Programming: vadd-gen}
    Generate efficient vector addition functions for given length.
    \begin{exampleblock}{}
    \begin{verbatim}
 val vadd-gen : Int -> (Vec -> Vec -> Vec) code
 # let vadd3 = vadd-gen 3
 val vadd3 : (Vec -> Vec -> Vec) code
 (* vadd3 v w =                            *)
 (*   < let (v, w)=(~v, ~w) in             *)
 (*      [v[0]+w[0];v[1]+w[1];v[2]+w[2]] > *)
    \end{verbatim}
    \end{exampleblock}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{The problem of vadd3 generated by vadd-gen}
    Runtime error occurs for vectors of wrong length.
    \begin{exampleblock}{}
    \begin{verbatim}
 val vadd3 : (Vec -> Vec -> Vec) code
 # (run vadd3) [1;2;3] [4;5;6]
 - : Vec = [5;7;9]
 # (run vadd3) [1;2] [4;5]
 Exception: Failure
    \end{verbatim}
    \end{exampleblock}
    \note{
    }
\end{frame}

\begin{frame}[fragile]{Dependent Types}
    \renewcommand{\V}{\text{Vec}\ }
    Types dependent on values
    \begin{itemize}
        \item Can prevent illegal use of {\verb|vadd3|}.
    \end{itemize}
    \begin{exampleblock}{}
 \begin{verbatim}
 [1;2]: Vec 2
 vadd3 : (Vec 3 -> Vec 3 -> Vec 3) code

 # (run vadd3) [1;2] [4;5]
 [1;2] has type Vec 2
 but was expected of type Vec 3
 \end{verbatim}
    \end{exampleblock}
    \note{
        By the way, we can write types dependent on values using dependent types.
    }
\end{frame}

@@

\begin{frame}[fragile]{\LTP [Hanada\&Igarashi'14]}
    \begin{table}
        \begin{tabular}{ c | c | c }
            & \LTP & MetaOCaml \\[2mm]
            \hline
            Quote & \( \textbf{let } a = \TB (1 + 2) \) & \verb| let a = <(1+2)> | \\[2mm]
            Unquote & \( (\TB (\TBL a + \TBL a)) \) & \verb| <(~a)+(~a)> | \\[2mm]
            Run & \( \textbf{run } (\TB (1 + 2)) \) & \verb| run <(1+2)> | \\[2mm]
            Code type & \( \TW\ \text{Int} \) & \verb| int code |
        \end{tabular}
    \end{table}
\end{frame}

\begin{frame}[fragile]{Terms of \LTP}
    \begin{block}{Terms}
        \begin{tabbing}
            \hspace{5mm} \( M \) \= \( ::= c \) \hspace{20mm} \= (constants) \\
            \> \( \mid \TB M \) \> (quote) \\
            \> \( \mid \TBL M \) \> (unquote) \\
            \> \( \mid \textbf{run}\ M \) \> (run) \\
            \> \( \mid \% M \) \> (Cross Stage Persistence (CSP)) \\
            \> \( \mid x \mid \lambda x:\tau.M \mid M\ M \)
        \end{tabbing}
    \end{block}
    * Simplified version
\end{frame}

\begin{frame}[fragile]{Types of \LTP}
    \begin{block}{Types}
        \begin{tabbing}
            \hspace{5mm} \( \tau \) \= ::= X \hspace{20mm} \= (type-level constants) \\
            \> \( \mid \TW\ \tau \) \> (code type) \\
        \> \( \mid \tau \to \tau \) \> (function type)
    \end{tabbing}
    \end{block}
   \note{
    }
\end{frame}

\begin{frame}[fragile]{What is a stage?}
    Level of nested quote
    \begin{exampleblock}{}
    \begin{center}
        \( \underbrace{(\textbf{run}\ (\TB\ \overbrace{(2 + 3)}^{\let\scriptstyle\textstyle\substack{\text{stage }1}})) + 1}_{\let\scriptstyle\textstyle\substack{\text{stage }0}} \)
    \end{center}
    \end{exampleblock}

@@

\begin{frame}[fragile]{Typing Rules of \LTP}
    Cannot apply a term to a function in a different stage.
    \begin{center}
        {\footnotesize \infrule[\TApp]{
            \G\V M:\sigma \to \tau @ n \andalso
            \G\V N:\sigma @ n
        }{
            \G\V M\ N : \tau @ n
        }}
    \end{center}
    Increase and decrease stage using \( \TB \) and \( \TBL \).
    \begin{center}
        {\footnotesize \infrule[{\TTB}]{
            \G\V M:\tau @ n+1
        }{
            \G\V\TB M:\TW \tau @ n
        } \hfil
        \infrule[{\TTBL}]{
            \G\V M:\TW \tau @ n
        }{
            \G\V\TBL M:\tau @ n + 1
        }}
    \end{center}
    In general, we cannot use terms of one stage at other stages.
    \note{
        Let's have a look on some typing rules.
    }
\end{frame}

@@

    Want to use \( \text{mul5} \) in a code value.
    \begin{exampleblock}{}
    \begin{center}
        \( \textbf{let}\ \underbrace{\text{mul5}}_{\text{stage }0}\ x = 5 * x\ \textbf{in} \TB\ \underbrace{(\% \text{mul5}\ 4)}_{\text{stage }1} \)
    \end{center}
    \end{exampleblock}
    CSP symbol \( \% \) {\bf{lift up}} mul5 from stage 0 to stage 1.
\end{frame}

@@

\begin{frame}[fragile]{Kinding rules of \LMD}
    Types have also stages because types may contains terms.
    \begin{center}
    {\footnotesize
    \infrule[{\KApp}]{
        \G\V \sigma:: (\Pi x:\tau.K) @ n \andalso \G\V M:\tau @ n
    }{
        \G\V \sigma\ M::K[x\mapsto M] @ n
    }
    \infrule[{\KAbs}]{
        \G\V \tau :: * @ n \andalso \G,x:\tau@ n \V \sigma::J@A
    }{
        \G\V(\Pi x:\tau.\sigma) :: (\Pi x:\tau.J) @ n
    }}
    \end{center}
    So, we need CSP in types also.
    \begin{exampleblock}{}
    \begin{center}
    \( \underbrace{\Pi n:\I. \TW (\overbrace{\text{Vec}\ \%n}^{\text{stage }1})}_{\text{stage }0} \)
    \end{center}
    \end{exampleblock}
\end{frame}

\begin{frame}[fragile]{Constants and type-level constants of \LMD}
    \begin{itemize}
        \item Constants: \( 0,1,2,3,\dots \)
        \item Type-level constants: Int, Vec, \dots
    \end{itemize}
    can be used at any stage.

    \begin{exampleblock}{}
        \begin{tabbing}
            \hspace{5mm} \= \text{val double-at-0} : \( \text{Vec }3 \to \text{Vec }3 \) \\[2mm]
            \> \textbf{let } \text{double} = \( \TB (\%\ \text{double-at-0}) \) \\
            \> \text{val double} : \( \TW \underbrace{(\text{Vec }3 \to \text{Vec }3)}_{\text{stage 1}} \)
        \end{tabbing}
    \end{exampleblock}
    Vec and integer 3 can be used at stage 1.
\end{frame}

\begin{frame}[fragile]{vadd-gen in \LMD}
    Type tells us the length of vector.
    \newcommand{\Vn}{\text{Vec}\ \%n}
    \newcommand{\Vt}{\text{Vec}\ \%3}
    \begin{exampleblock}{}
    \begin{tabbing}
        \hspace{5mm} \= \( \text{vadd-gen} : \Pi n:\I. \TW (\Vn \to \Vn \to \Vn) \) \\[2mm]
        \> \( \textbf{let}\ \text{vadd3} = \text{vadd-gen}\ 3;; \) \\
        \> \( \text{val}\ \text{vadd3} : \TW (\Vt \to \Vt \to \Vt) \) \\[2mm]
    \end{tabbing}
    \end{exampleblock}
    Now, vadd3 takes only vectors of length 3.
    \note{
        vadd-gen is changed as follows in \LMD.
        Thanks to dependent types ...
    }
\end{frame}

\begin{frame}[fragile]{Technical challenge of vadd3}
    Hard to compose with other functions.
    \newcommand{\Vn}{\text{Vec}\ \%n}
    \newcommand{\Vpt}{\text{Vec}\ \%3}
    \newcommand{\Vt}{\text{Vec}\ 3}
    \begin{exampleblock}{}
    \begin{tabbing}
        \hspace{5mm} \= \( \text{val}\ \text{vadd3} : \TW (\Vpt \to \Vpt \to \Vpt) \) \\
        \> \( \text{val}\ \text{double} : \TW (\Vt \to \Vt) \) \\[2mm]
        \> \( \textbf{let}\ \text{add-double}\ v\ w = \) \\
        \> \hspace{5mm} \( \TB (\underbrace{\TBL\text{double}}_{\Vt \to \Vt}\ \underbrace{(\TBL \text{vadd3}\ \TBL v\ \TBL w)}_{\Vpt});; \) \\
    \end{tabbing}
    \end{exampleblock}
    We cannot compose because of type mismatch.
\end{frame}

\begin{frame}[fragile]{Solution}
    \newcommand{\Vt}{\text{Vec}\ 3}
    \newcommand{\Vpt}{\text{Vec}\ \%3}
    Erase the CSP symbol \( \% \) of an closed term.
    \begin{center}
        \infrule[{\QPercent}]{
            M \text{ is a closed term.}
        }{
            \G\vdash \% M \E M
        }
    \end{center}
    \begin{exampleblock}{}
    \begin{itemize}
        \item \( \text{Vec}\ \%3 \E \text{Vec}\ 3 \)
        \item \( \text{vadd3} : \TW (\Vt \to \Vt \to \Vt) \) \\
            \hspace{15mm} \( (\E  \TW (\Vpt \to \Vpt \to \Vpt)) \)
    \end{itemize}
    \end{exampleblock}
\end{frame}
